Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-06-10 GWLDCI.tex
fixed the rest of the overwritten changes

\sanote{I undid Sian's edit here. I like "saying what i'm about to say. Saying it. Summarizing what I said"}In this section we have characterized the correspondence between Wilson loop diagrams and positroids, showing that Wilson Loop diagrams define the same matroid if and only if they are equivalent (Theorem \ref{same matroid iff equiv}). Corollary \ref{number of equiv diagrams} enumerates the fibres of the map from Wilson Loop diagrams to positroids.  The obvious next question would be to enumerate the number of equivlence classes.  We do not obtain a formula for this, but do obtain an alternate characterization of the equvalence classes which is the subject of the next section.

\section{Associahedra and inequivalent Wilson loop diagrams}\label{sec associahedron}

The connection between admissible Wilson loop diagrams and polytope dissections suggests a connection to associahedra. First we need some background. \sanote{expand}

\subsection{Polytope background}\label{sec polytope background}
We first give a quick summary of the polytope terminology we will use in this section; the interested reader is referred to \cite{Ziegler} for more details.

@@

A \emph{face} of a polytope is the intersection of the polytope with any hyperplane for which the entire polytope is in one of the closed half-spaces defined by the hyperplane.  The polytope itself is also considered a face (in the case that that polytope is of full dimension, this definition can be unified by considering linear equalities (including the trivial one) in place of hyperplanes and the corresponding linear inequalities in place of the half-spaces defined by the hyperplane, see chapter 2 of \cite{Ziegler}).  Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.

Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.  In particular parallel faces must be of the same dimension and any two vertices are parallel to each other. There are many important polytopes in combinatorics, including, of primary importance to us, the associahedron.

Finally, two polytopes are \emph{combinatorially equivalent} if there is a bijection between their faces which preserves inclusion of faces.  Often in combinatorics one is primarily interested in the combinatorial equivalence classes, and common names, including the associahedron, refer to these equivalence classes.  A particular polytope in the class is then known as a \emph{realization}.

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}; the following will be the one which we will use.  See Figure~\ref{associahedron5gon} for examples.

\begin{dfn} \label{def:secondary polytope}
  Let $x_1, \ldots, x_n$ define a convex $n$-gon and let $T$ be the set of triangulations of this $n$-gon.  For each $t\in T$ define $s_t$ to be the point in $\mathbb{R}^n$ whose $i$th coordinate is the sum of the areas of all triangles of $t$ which are incident to $x_i$.  Let $A_n$ be the convex hull of the $s_t$; then $A_n$ a realization of the $n-1$ associahedron (\cite{Ziegler}, Example 9.11)
\end{dfn}

Note that this construction works for any convex $n$-gon, but these realizations are all equally good for our purposes, and so we fix $x_1, \ldots, x_n$ defining a particular convex $n$-gon from now on.

Definition~\ref{def:secondary polytope} is a special case of the {\em secondary polytope} construction; see \cite[Definition 9.9]{Ziegler} for more details.  In general the $x_1, \ldots, x_n$ may be any points in any $\mathbb{R}^d$.  Triangulations in this more general context are decompositions into simplices and the area of the triangles is replaced by the volume of the simplices.  
This is the construction of the secondary polytope (so $A_n$ is the secondary polytope of $x_1, \ldots, x_n$), see \cite{Ziegler} Definition 9.9.  In general the $x_1, \ldots, x_n$ may be any points in any $\mathbb{R}^d$.  Triangulations in this more general context are decompositions into simplices and the area of the triangles is replaced by the volume of the simplices.

@@

Note that Wilson loop diagrams which are weakly admissible but not admissible are necessarily exact, and hence have $\tau$ graphs which are triangulations, i.e. dimension 0 faces (vertices) in $A_n$.  All of these are equivalent (as Wilson loop diagrams) in the sense that they are only retriangulations away from each other; meanwhile, as noted above, all vertices of $A_n$ are parallel to each other.  It will be convenient in what follows to include these diagrams, but the results are easily modified to remove them if needed: Propositions~\ref{prop easy way} and \ref{prop hard way} would be identical if we replaced ``weakly admissible'' by ``admissible'', while Corollary~\ref{cor:count inequiv diagrams} could be changed to consider only admissible diagrams by subtracting 1 from the number of non-parallel faces in order to account for the missing dimension 0 faces.Note that Wilson loop diagrams which are weakly admissible but not admissible are necessarily exact, and hence have $\tau$ graphs which are triangulations, i.e. dimension 0 faces (vertices) in $A_n$.  All of these are equivalent (as Wilson loop diagrams) in the sense that they are only retriangulations away from each other; meanwhile, as noted above, all vertices of $A_n$ are parallel to each other.  It will be convenient in what follows to include these diagrams, but the results are easily modified to remove them if needed: Propositions~\ref{prop easy way} and \ref{prop hard way} would be identical if we replaced ``weakly admissible'' by ``admissible'', while Corollary~\ref{cor:count inequiv diagrams} could be changed to consider only admissible diagrams by subtracting 1 from the number of non-parallel faces in order to account for the missing dimension 0 faces.

@@

Consider now an edge in $A_n$ whose quadrilateral is defined by the vertices of $\tau(W)$ or $\tau(X), $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From Definition \ref{def:secondary polytope}, defining $A_n$,  we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$. All of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates.  This is illustrated in Figure~\ref{triangulate}.

@@

    Proceed now to check that $\omega^+_d$ is a normal vector for any face associated to a dissection including $d$.  We will do this by showing it is normal to every edge bounding the face.  To do so, choose the coordinate system for the polygon so that $d$ is along the $x$-axis.  Let $(a_i, h_i)$, $i=1,2,3,4$ be the coordinates of the vertices of a quadrilateral in counterclockwise order and which corresponds to an edge bounding the face.  Since the dissection includes $d$ either all four points are on the negative side of $d$, and so the dot product of $\omega^+_d$ with the direction vector for the quadrilateral is immediately $0$, or, by the formula in the proof of the previous proposition and the shoelace formula, the dot product of $\omega^+_d$ with the vector associated to the quadrilateral is
    \begin{align*}
      & h_1(a_2h_3-a_2h_4+a_3h_4-a_3h_2+a_4h_2-a_4h_3)/2 - h_2(a_1h_3-a_1h_4+a_3h_4-a_3h_1+a_4h_1-a_4h_3)/2 \\
      & + h_3(a_1h_2-a_1h_4+a_2h_4-a_2h_1+a_4h_1-a_4h_2)/2 - h_4(a_1h_2-a_1h_3+a_2h_3-a_2h_1+a_3h_1-a_3h_2)/2 \\
      & = 0
    \end{align*}

>>>>

\sanote{I undid Sian's edit here. I like "saying what i'm about to say. Saying it. Summarizing what I said"}\note{Sian: sure. I'll push back if it gets too repetitive, but it works here.}In this section we have characterized the correspondence between Wilson loop diagrams and positroids, showing that Wilson Loop diagrams define the same matroid if and only if they are equivalent (Theorem \ref{same matroid iff equiv}). Corollary \ref{number of equiv diagrams} enumerates the fibres of the map from Wilson Loop diagrams to positroids.  The obvious next question would be to enumerate the number of equivlence classes.  We do not obtain a formula for this, but do obtain an alternate characterization of the equvalence classes which is the subject of the next section.

\subsection{Polytope background}\label{sec polytope background}

We first give a quick summary of the polytope terminology we will use in this section; the interested reader is referred to \cite{Ziegler} for more details.

@@

A \emph{face} of a polytope is the intersection of the polytope with any hyperplane for which the entire polytope is in one of the closed half-spaces defined by the hyperplane.  The polytope itself is also considered a face; in the case that that polytope is of full dimension, this definition can be unified by considering linear equalities (including the trivial one) in place of hyperplanes and the corresponding linear inequalities in place of the half-spaces defined by the hyperplane (see chapter 2 of \cite{Ziegler}).

Two faces are \emph{parallel} if one can be mapped to the other by a translation in $\mathbb{R}^d$.  In particular parallel faces must be of the same dimension and any two vertices are parallel to each other.

Finally, two polytopes are \emph{combinatorially equivalent} if there is a bijection between their faces which preserves inclusion of faces.  Often in combinatorics one is primarily interested in the combinatorial equivalence classes, and common polytope names (including the associahedron) refer to these equivalence classes.  A particular polytope in the class is then known as a \emph{realization}.

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}; the following realization will be the one which we will use.

\begin{dfn}\label{def:secondary polytope}
  Let $x_1, \ldots, x_n$ define a convex $n$-gon, and let $T$ be the set of triangulations of this $n$-gon.  For each $t\in T$, define $s_t$ to be the point in $\mathbb{R}^n$ whose $i$th coordinate is the sum of the areas of all triangles of $t$ which are incident to $x_i$.  Let $A_n$ be the convex hull of the $s_t$; then $A_n$ a realization of the $n-1$ associahedron (\cite[Example 9.11]{Ziegler}).
\end{dfn}
Note that this construction works for any convex $n$-gon, but these realizations are all equally good for our purposes, and so we fix $x_1, \ldots, x_n$ defining a particular convex $n$-gon from now on.

Definition~\ref{def:secondary polytope} is a special case of the {\em secondary polytope} construction; see \cite[Definition 9.9]{Ziegler} for more details.  In general the $x_1, \ldots, x_n$ may be any points in any $\mathbb{R}^d$.  Triangulations in this more general context are decompositions into simplices and the area of the triangles is replaced by the volume of the simplices.

@@

Note that Wilson loop diagrams which are weakly admissible but not admissible are necessarily exact, and hence have $\tau$ graphs which are triangulations, i.e. dimension 0 faces (vertices) in $A_n$.  All of these are equivalent (as Wilson loop diagrams) in the sense that they are only retriangulations away from each other; meanwhile, as noted above, all vertices of $A_n$ are parallel to each other.  It will be convenient in what follows to include these diagrams, but the results are easily modified to remove them if needed: Propositions~\ref{prop easy way} and \ref{prop hard way} would be identical if we replaced ``weakly admissible'' by ``admissible'', while Corollary~\ref{cor:count inequiv diagrams} could be changed to consider only admissible diagrams by subtracting 1 from the number of non-parallel faces in order to account for the missing dimension 0 faces.

@@

Consider now an edge in $A_n$ whose quadrilateral is defined by the vertices of $\tau(W)$ or $\tau(X)$, $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From Definition \ref{def:secondary polytope}, defining $A_n$,  we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$. All of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates.  This is illustrated in Figure~\ref{triangulate}.

@@

    Proceed now to check that $\omega^+_d$ is a normal vector for any face associated to a dissection including $d$.  We will do this by showing it is normal to every edge bounding the face.  To do so, choose the coordinate system for the polygon so that $d$ is along the $x$-axis.  Let $(a_i, h_i)$, $i=1,2,3,4$ be the coordinates of the vertices of a quadrilateral in counterclockwise order and which corresponds to an edge bounding the face.  Since the dissection includes $d$ either all four points are on the negative side of $d$, and so the dot product of $\omega^+_d$ with the direction vector for the quadrilateral is immediately $0$, or, by the formula in the proof of the previous proposition and the shoelace formula, the dot product of $\omega^+_d$ with the vector associated to the quadrilateral is
   \begin{align*}
       & \hspace{1cm}\textstyle\frac{1}{2}h_1(a_2h_3-a_2h_4+a_3h_4-a_3h_2+a_4h_2-a_4h_3) \\
       &\hspace{3cm} - \textstyle\frac{1}{2} h_2(a_1h_3-a_1h_4+a_3h_4-a_3h_1+a_4h_1-a_4h_3) \\
       &\hspace{5cm}+ \textstyle\frac{1}{2}h_3(a_1h_2-a_1h_4+a_2h_4-a_2h_1+a_4h_1-a_4h_2) \\
       &\hspace{7cm}- \textstyle\frac{1}{2}h_4(a_1h_2-a_1h_3+a_2h_3-a_2h_1+a_3h_1-a_3h_2), 
    \end{align*}
    which evaluates to 0.
